---
title: "Diamonds_Problem_Set"
output:
  html_document: default
  pdf_document: default
---
##Problem Set 1 - Exploring 1 Variable

###Exploring Diamond Price Distribution
```{r, echo=FALSE}

library(ggplot2)
library(gridExtra)
library(dplyr)
library(tidyr)
data(diamonds)
str(diamonds)
table(diamonds$color)

p1 <- ggplot(aes(x=diamonds$price),data=diamonds) + geom_histogram() + scale_x_continuous(breaks=seq(0,10000,1000),limits=c(0,10000)) + theme_classic() + xlab('Price') + ylab('Number of Diamonds')

p2 <- ggplot(aes(x=diamonds$price),data=diamonds) + geom_histogram() + scale_x_continuous(breaks=seq(300,1500,100),limits=c(300,1500)) + theme_classic() + xlab('Price') + ylab('Number of Diamonds')

grid.arrange(p1,p2)

length(diamonds$price[diamonds$price<500])
length(diamonds$price[diamonds$price<250])
length(diamonds$price[diamonds$price>=15000])

ggsave('diamonds.jpeg')

```

###Exploring Diamond Price Distribution by Cut
```{r, echo=FALSE}

by(diamonds$price,diamonds$cut,summary,digits = max(getOption('digits')))

ggplot(aes(x=diamonds$price),data=diamonds) + facet_wrap(~cut,scales='free')  + xlab('Price') + ylab('Number of Diamonds') + geom_histogram()

ggsave('diamonds_cut.jpeg')
```

###Exploring Diamond Price/Carat Distribution by Cut
````{r, echo=FALSE}

by(log(diamonds$price/diamonds$carat),diamonds$cut,summary,digits = max(getOption('digits')))

ppc <- ggplot(aes(x=diamonds$price/diamonds$carat),data=diamonds) + facet_wrap(~cut,scale='free')  + xlab('Price per Carat') + ylab('Number of Diamonds') + geom_histogram() + theme_classic()

ppc + scale_x_log10() + xlab('Log Price per Carat') 
````

###Exploring Diamond Price and Price/Carat Distribution by Color
````{r, echo=FALSE}

by(log(diamonds$price/diamonds$carat),diamonds$color,summary,digits = max(getOption('digits')))

pc <- ggplot(aes(x=diamonds$price),data=diamonds) + facet_wrap(~color,scales='free')  + ylab('Number of Diamonds') + xlab(NULL) + geom_histogram() + theme_classic() + scale_x_log10() + ggtitle('Log Price by Color')

ppcc <- ggplot(aes(x=diamonds$price/diamonds$carat),data=diamonds) + facet_wrap(~color,scales='free')  +  ylab('Number of Diamonds') + xlab(NULL) + geom_histogram() + theme_classic() + scale_x_log10() + ggtitle('Log Price per Carat by Color')

grid.arrange(pc,ppcc,ncol=2)

pb <- ggplot(data = diamonds,aes(x=color,y=log10(price))) + geom_boxplot() + theme_classic() + ylab('Log Price') + ggtitle('Log Price by Color')

pcb <- ggplot(data = diamonds,aes(x=color,y=log10(price/carat))) + geom_boxplot() + theme_classic() + ylab('Log Price per Carat') + ggtitle('Log Price per Carat by Color')

grid.arrange(pb,pcb,ncol=2)

table(diamonds$color)
by(diamonds$carat,diamonds$color,mean)

pcbar <- ggplot(data = diamonds,aes(x=color)) + geom_bar() + theme_classic() + ylab('Number of Diamonds') + xlab('Color') + ggtitle('Frequency of Each Color')

diamonds$carat_cat <- cut(diamonds$carat,breaks=c(0,.5,1,1.5,2,2.5,3,6))

ccbar <- ggplot(data = diamonds,aes(x=carat_cat)) + geom_bar() + theme_classic() + ylab('Number of Diamonds') + xlab('Number of Carats') + ggtitle('Frequency of Carat Categories')

table(diamonds$carat_cat,diamonds$color)
nrow(subset(diamonds,diamonds$carat<2))/nrow(diamonds)*100
#96% of diamonds have fewer than 2 carats, so we'll get good results looking only at these diamonds

nrow(subset(diamonds,diamonds$color=='D'))
D_diamonds <- subset(diamonds,diamonds$color=='D')
table(D_diamonds$carat_cat)
#D diamonds are uncommon, but seem to be distributed across the carat categories in a similar way to the dataset overall--this isn't the source of the skew

high_carat_diamonds <- (subset(diamonds,diamonds$carat>2))
table(high_carat_diamonds$color)

color_cat <- ggplot(data = subset(diamonds,diamonds$carat<2),aes(x=color,y=log10(price))) + facet_wrap(~carat_cat) + geom_boxplot() + theme_minimal() + ylab('Log Price') + ggtitle('Log Price by Color and Number of Carats') + xlab(NULL)

ggsave('price_by_color_and_carats.jpeg')

````

###Interquartile range
```{r,echo=FALSE}

quantile(subset(diamonds,diamonds$color=='D')$price,seq(0,1,.25))
quantile(subset(diamonds,diamonds$color=='J')$price,seq(0,1,.25))
IQR(subset(diamonds,diamonds$color=='D')$price)
IQR(subset(diamonds,diamonds$color=='J')$price)

```

###Price per carat by color boxplot
```{r,echo=FALSE}
color_box <- ggplot(data = diamonds,aes(x=color,y=price/carat)) + geom_boxplot() + theme_minimal() + ylab('Price per Carat') + ggtitle('Price per Carat by Color') + xlab(NULL)

ggsave('ppc_by_color.jpeg')
```

###Weight distribution
```{r,echo=FALSE}

f_carat <- ggplot(aes(x=carat),data=diamonds) + xlab('Number of Carats') + ylab('Count of Diamonds') + scale_x_continuous(breaks=seq(0,3,.1),limits=c(0,2)) + scale_y_continuous(breaks=seq(0,10000,1000)) + geom_freqpoly() + theme_classic()


```



##Problem Set 2 - Exploring 2 Variables
###Price and Volume
```{r,echo=FALSE}

#price vs. x, y, and z
price_v_x <- ggplot(data=diamonds,aes(x=x,y=price)) + 
  scale_x_continuous(breaks=seq(0,15,1)) +
  geom_point(position=position_jitter()) +
  ggtitle('price and x') +
  geom_smooth() +
  theme_classic()

standardized_price <- with(diamonds,price/sd(price))
sprice <- qplot(standardized_price) + theme_classic()
standardized_x <- with(diamonds,x/sd(x))
sx <- qplot(standardized_x) + theme_classic()
standardized <- grid.arrange(sprice,sx,ncol=2)

grid.arrange(price_v_x,standardized,nrow=2)

cor(diamonds$x,diamonds$price)
cor(diamonds$y,diamonds$price)
cor(diamonds$z,diamonds$price)

#price vs. x, y, and z
# install.packages('corrplot')
# library(corrplot)
# D <- cor(diamonds[,-(2:4)])
# corrplot(D,'circle','lower')

price_v_depth <- ggplot(data=diamonds,aes(x=depth,y=price)) + 
  geom_point(position=position_jitter(),alpha=1/100) +
  ggtitle('price and depth') +
  scale_x_continuous(breaks=seq(0,max(diamonds$depth),2)) +
  theme_classic()

price_v_depth
quantile(diamonds$depth)
cor(diamonds$depth,diamonds$price)

ggplot(data=subset(diamonds,diamonds$price<quantile(diamonds$price,.99)),
       aes(x=carat,y=price)) +
  geom_point(alpha=1/15,position=position_jitter()) +
  scale_x_continuous(breaks=seq(0,5,.5))

#price vs volume
diamonds$volume <- with(diamonds,x*y*z)

ggplot(data=diamonds,
       aes(x=volume,y=price)) +
  geom_point()

diamonds_no_outliers <- subset(diamonds,
                               diamonds$volume>0&
                                 diamonds$volume<=800&
                                 diamonds$price>0)

cor(diamonds_no_outliers$price,diamonds_no_outliers$volume)

ggplot(data=diamonds_no_outliers,
       aes(x=volume,y=price)) +
  geom_point(alpha=1/25) +
  scale_y_log10() +
  geom_smooth(formula=log10(y)~x,colour='red') +
  coord_trans(limx=c(0,500)) +
  ggtitle('volume and log of price') +
  theme_bw()
  
#New dataframe: diamonds by clarity

diamonds_no_outliers %>% group_by(carat) %>% summarise(mean_price = mean(price), median_price = median(price), min_price = min(price), max_price = max(price), n())

  #tangent: practice creating a matrix of values for a given mean, median combination to easily plot different summary stats
  diamonds_melted <- diamonds_no_outliers %>% group_by(carat) %>% summarise(mean_price = mean(price), median_price = median(price), min_price = min(price), max_price = max(price), n = n()) %>% gather('metric','value',c('mean_price','median_price'))

  diamonds_mean_median <- ggplot(data=diamonds_melted,aes(x=carat,y=metric)) + geom_line(aes(x=carat,y=value,colour=metric)) + ylab('price')

```

```{r,echo=FALSE}

library(dplyr)

diamonds_by_clarity <- group_by(diamonds, clarity)
diamonds_mp_by_clarity <- summarise(diamonds_by_clarity, mean_price = mean(price))

diamonds_by_color <- group_by(diamonds, color)
diamonds_mp_by_color <- summarise(diamonds_by_color, mean_price = mean(price))

diamonds_mp_by_cut <- diamonds %>% group_by(cut) %>% summarise(mean_price = mean(price))

clar <- ggplot(diamonds_mp_by_clarity,aes(x=clarity,y=mean_price)) +
  geom_bar(stat='identity') +
  ylab('mean price') +
  theme_bw()

col <- ggplot(diamonds_mp_by_color,aes(x=color,y=mean_price)) +
  geom_bar(stat='identity') +
  ylab('mean price') +
  theme_bw()

cut <- ggplot(diamonds_mp_by_cut,aes(x=cut,y=mean_price)) +
  geom_bar(stat='identity') +
  ylab('mean price') +
  theme_bw()

ggsave(plot=grid.arrange(clar,col,cut),file='diamond price by color, clarity, and cut','jpeg')

grid.arrange(clar,col,cut)

```




##Problem Set 3 - Exploring 3 Variables

###Price by Color and Cut
```{r,echo=FALSE}

ggplot(data=diamonds,aes(x=price)) +
  geom_histogram(aes(fill=cut),binwidth=.1) +
  scale_x_log10() +
  facet_wrap(~color,scales='free') +
  theme_minimal() +
  scale_fill_brewer(palette='PuBuGn') +
  ylab('frequency') +
  ggtitle('log diamond price by cut and clarity')

```

###Price by Table and Cut
```{r,echo=FALSE}

plot <- ggplot(data=diamonds,aes(x=table,y=price)) +
  geom_point(aes(colour=cut),position=position_jitter(),alpha=1/15) +
  theme_minimal() +
  guides(colour = guide_legend(override.aes = list(alpha=1))) +
  ggtitle('diamond price by table and cut') +
  scale_x_continuous(breaks=seq(0,100,2)) +
  coord_trans(limx=c(40,80))

plot
by(diamonds$table,diamonds$cut,summary)


```

###Price by Volume and Clarity
```{r,echo=FALSE}

ggplot(data=subset(diamonds,volume>0&volume<quantile(diamonds$volume,.99)),aes(x=volume,y=price,colour=clarity)) +
  geom_point(alpha=1/20) +
  scale_y_log10() +
  guides(colour=guide_legend(override.aes = list(alpha=1))) +
  theme_minimal() +
  scale_colour_brewer(palette='PuBuGn') +
  ggtitle('Log Price vs. Volume and Clarity')

```

###Price by Volume and Clarity

Create a scatter plot of the price/carat ratio of diamonds. 
The variable x should be assigned to cut. The points should be colored
by diamond color, and the plot should be faceted by clarity.

```{r,echo=FALSE}

ggplot(data=diamonds,aes(x=cut,y=price/carat,colour=color)) +
  geom_point(alpha=1/10,position=position_jitter()) +
  facet_wrap(~clarity) +
  guides(colour=guide_legend(override.aes = list(alpha=1))) +
  scale_colour_brewer(palette='BuGn') +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90))

```


